---
title: "DATA 607 - Assignment 5A - Approach"
author: "Kailot C. Harris"
date: "2026/09/25"
format: html
editor: visual
theme:
  dark: darkly
  light: flatly
---

# Introduction and Overview

Data is provided in the form of a formatted table with frequencies of on-time and delayed arrivals. An initial goal of this work is to accurately store this information in a .CSV file. Next, the .CSV file will be read into R using the `readr` package's `read_csv()` function. The `tidyr` and `dplyr` packages will be utilized to tidy and transform data. Arrival delay information will be compared across airlines and visualized.

# Codebase

## Data Import

```{r read-csv}
library(tidyverse)

url <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Assignment-5A/refs/heads/main/numberverse.csv"

csv_data <- readr::read_csv(file = url, 
                            show_col_types = FALSE,
                            progress = TRUE)
```

## Data Tidying

### Pivoting Longer

An initial step in data tidying is pivoting the data frame longer because important data are contained in column names (R 4 Data Science, Section 5.3.1). By converting each destination city's column into frequency data and making a new column for destination city names, the data is prepared for further analysis.

```{r data-tidying}
df_long <- csv_data |>
  pivot_longer(
    cols = c(3,4,5,6,7),
    names_to = "destination",
    values_to = "frequency"
  )
head(df_long)
```

### Summarizing Data

To prepare the data for summarizing, the data frame was pivoted wider. This replaced the

```{r data-summary}

df_long <- df_long |>
  group_by(status) |>
  arrange(destination)

  
df_wide <- df_long |>
  pivot_wider(
    names_from = status,
    values_from = frequency
  )

df_wide <- janitor::clean_names(df_wide)
  
df_summary <- df_wide |>
  mutate(on_time_pct = round(100*on_time/(on_time+delayed),1)) |>
  select(airline, destination, on_time_pct)

df_wider <- df_summary |>
  pivot_wider(names_from = airline,
              values_from = on_time_pct
  ) |>
  janitor::clean_names() |>
  arrange(desc(alaska))


knitr::kable(df_wider, caption = "Summary of Data")
```

## Visualizing Data

```{r viz-1}
df_wider |>
  ggplot(aes(x=destination, y=alaska)) +
  geom_col() +
  scale_y_continuous(limits = c(0,100)) +
  labs(x="Destination",
       y="Percent On Time") +
  theme_minimal()



```
