---
title: "DATA 607 - Assignment 5A - Approach"
author: "Kailot C. Harris"
date: "2026/09/25"
format: html
editor: visual
theme:
  dark: darkly
  light: flatly
---

# Introduction and Overview

Data is provided in the form of a formatted table with frequencies of on-time and delayed arrivals. An initial goal of this work is to accurately store this information in a .CSV file. Next, the .CSV file will be read into R using the `readr` package's `read_csv()` function. The `tidyr` and `dplyr` packages will be utilized to tidy and transform data. Arrival delay information will be compared across airlines and visualized.

# Codebase

## Data Import

```{r read-csv}
library(tidyverse)

url <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Assignment-5A/refs/heads/main/numberverse.csv"

csv_data <- readr::read_csv(file = url, 
                            show_col_types = FALSE,
                            progress = TRUE)
```

## Data Tidying

### Pivoting Longer

An initial step in data tidying is pivoting the data frame longer because important data are contained in column names (R 4 Data Science, Section 5.3.1). By converting each destination city's column into frequency data and making a new column for destination city names, the data is prepared for further analysis.

```{r data-tidying}
df_long <- csv_data |>
  pivot_longer(
    cols = c(3,4,5,6,7),
    names_to = "destination",
    values_to = "frequency"
  )
head(df_long)
```

### Summarizing Data

To prepare the data for summarizing, the data frame was pivoted wider. This replaced the frequency column with two columns per unique combination of `airline` & `destination`: `on_time` and `delayed`. From there a new column, `on_time_pct`, was mutated in to calculate the percent of on-time flights for each unique combination of `airline` & `destination` . The data frame was then pivoted even wider, generating unique columns for `alaska` and `am west` airlines with their respective `on_time_pct` calculations.

```{r data-summary}

df_long <- df_long |>
  group_by(status) |>
  arrange(destination)

df_wide <- df_long |>
  pivot_wider(
    names_from = status,
    values_from = frequency
  )

df_wide <- janitor::clean_names(df_wide)
  
df_summary <- df_wide |>
  mutate(on_time_pct = round(100*on_time/(on_time+delayed),1)) |>
  select(airline, destination, on_time_pct)

df_wider <- df_summary |>
  pivot_wider(names_from = airline,
              values_from = on_time_pct
  ) |>
  janitor::clean_names() |>
  arrange(desc(alaska))


knitr::kable(df_wider, caption = "Summary of Data")
```

## Visualizing Data

A column plot compares the `on_time_pct` values for each

```{r viz-1}
library(patchwork)
df_summary |>
  ggplot(aes(x=destination, y=on_time_pct, fill=destination)) +
  facet_wrap(~airline) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(limits = c(0,100)) +
  labs(title = "Comparing Arrival Delays Across Airline and Destination",
       x="Destination",
       y="Percent On Time") +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 13, face = "italic"),
    axis.text.y = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(size = 11, face = "bold", angle = 45, hjust = 0.8, vjust = 1.1),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
  ) +
  scale_fill_discrete()
```

```{r viz-2}


```
