---
title: "DATA 607 - Assignment 5A - Codebase"
author: "Kailot C. Harris"
date: "2026/09/27"
format:
  html:
    code-overflow: wrap
editor: visual
theme:
  dark: darkly
  light: flatly
---

# Introduction and Overview

Data is provided in the form of a formatted table with frequencies of on-time and delayed arrivals. An initial goal of this work is to accurately store this information in a .CSV file. Next, the .CSV file will be read into R using the `readr` package's `read_csv()` function. The `tidyr` and `dplyr` packages will be utilized to tidy and transform data. Arrival delay information will be compared across airlines and visualized.

Anticipated challenges in this assignment include working with missing data, tidying and transforming the data frame appropriately, and drawing comparisons between airlines and destinations. This work will endeavor to answer whether the on-time-rate winner changes between a city-by-city comparison and an overall comparison.

# Codebase

## Data Import

Data is fetched from a github-hosted .CSV file containing the relevant data. Next, missing airline data is imputed using the `fill()` function from within the `tidyverse`. The results of the `fill()` operation were manually inspected for correctness.

```{r read-csv}
#| echo: true
#| warning: false
#| message: false

library(tidyverse)

url <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Assignment-5A/refs/heads/main/numberverse.csv"

csv_data <- readr::read_csv(file = url, 
                            show_col_types = FALSE,
                            progress = FALSE)

clean_data <- csv_data |>
  fill(airline, .direction = "down")

```

## Data Tidying

### Pivoting Longer

An initial step in data tidying is pivoting the data frame longer because important data are contained in column names (R 4 Data Science, Section 5.3.1). By converting each destination city's column into frequency data and making a new column for destination city names, the data is prepared for further analysis.

```{r data-tidying}
#| echo: true
#| warning: false
#| message: false

df_long <- clean_data |>
  pivot_longer(
    cols = !c(airline, status),
    names_to = "destination",
    values_to = "frequency"
  )
```

### Summarizing Data

To prepare the data for summarizing, the data frame was pivoted wider. This replaced the frequency column with two columns per unique combination of `airline` & `destination`: `on_time` and `delayed`. From there a new column, `on_time_pct`, was mutated in to calculate the percent of on-time flights for each unique combination of `airline` & `destination` . The data frame was then pivoted even wider, generating unique columns for `alaska` and `am west` airlines with their respective `on_time_pct` calculations.

```{r data-summary}
#| echo: true
#| warning: false
#| message: false

df_wide <- df_long |>
  pivot_wider(
    names_from = status,
    values_from = frequency
  ) |>
  janitor::clean_names()

df_summary <- df_wide |>
  mutate(on_time_pct = round(100*on_time/(on_time+delayed),1)) |>
  select(airline, destination, on_time_pct)

df_wider <- df_summary |>
  pivot_wider(names_from = airline,
              values_from = on_time_pct
  ) |>
  janitor::clean_names() |>
  arrange(desc(alaska))


knitr::kable(df_wider, caption = "Summary of Data")
```

## Visualizing Data

```{r viz-1}
#| echo: true
#| warning: false
#| message: false
#| label: fig-1
#| fig-cap: "Comparing On-Time Rates Across Airline and Destination"

theme_standard <- function(){
  theme_minimal() +
    theme(
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.title = element_text(size = 15, face = "bold"),
      plot.subtitle = element_text(size = 13, face = "italic"),
      axis.text.y = element_text(size = 11, face = "bold"),
      axis.text.x = element_text(size = 11, face = "bold"),
      axis.title.y = element_text(size = 12, face = "bold"),
      axis.title.x = element_text(size = 12, face = "bold")
    )
}

df_summary |>
  ggplot(aes(x=destination, y=on_time_pct, fill=destination)) +
  facet_wrap(~airline) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(limits = c(0,100)) +
  labs(title = "Comparing On-Time Rates Across Airline and Destination",
       x="Destination",
       y="Percent On Time") +
  theme_standard()
```

```{r viz-2}
#| echo: true
#| warning: false
#| message: false
#| label: fig-2
#| fig-cap: "Alaska Airlines Displays a Higher On-Time Percentage than Am West"

df_viz <- df_wider |>
  mutate(diff = alaska - am_west) |>
  select(destination, diff)

df_viz |>
  ggplot(aes(x=destination, y=diff, fill=destination)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Alaska Airlines Displays a Higher On-Time Percentage than Am West",
       subtitle = "",
       x = "Destination",
       y = "Differential in Percent On-Time\n (Alaska minus Am West)"
       ) +
  coord_flip() +
  theme_standard()
```

To view the aggregate level summaries, the `df_wide` data frame is reused and summary statistics are summarized into the new frame, `df_sum`. From there, the percent of on time flights for each airline, aggregated across destinations, is computed.

```{r summary-2}
#| echo: true
#| warning: false
#| message: false
#| label: tbl-3
#| tbl-cap: "Aggregate Summary"

df_sum <- df_wide |>
  group_by(airline) |>
  summarize(
    tot_on_time = sum(on_time),
    tot_delayed = sum(delayed),
    pct_on_time = round(100*tot_on_time / (tot_on_time+tot_delayed),1)
  ) |>
  select(airline, pct_on_time)

knitr::kable(df_sum)

```

# Findings and Recommendations

In this assignment a .CSV file was loaded into a data frame using `read_csv()` and the data frame was tidied and reorganized in several iterations. The frame was transformed wider and longer from its initial shape. In the end, the data is described in two visualizations comparing on-time rates (percentage) for five destinations across two airlines and the difference in on-time rates for the two airlines.

According to @fig-1, each airline shows on-time rates exceeding 70%, with data from all of the Alaska airline destinations exceeding 80%. Flights to Phoenix represented the highest on-time rates for both airlines, while San Francisco represented the destination with the lowest on-time rates.

In @fig-2, it is clear that Alaska airline has higher on-time percent rates than Am West airline, when the data is stratified by destination. Here, the differential for each destination (calculated as Alaska on-time rate minus Am West on-time rate) is shown, which compares the on-time rates for the airlines across all five destinations. This chart makes apparent that the largest differential in on-time rates occurs when San Francisco is the destination.

Finally, a comparison of the two airlines on aggregate statistics is presented in @tbl-3. This table exposes a discrepancy in the data: in @fig-1 and @fig-2 it was clear that Alaska airline outperformed Am West airline when results were dis-aggregated by destination; however, in @tbl-3, it is shown that Am West outperforms Alaska airline in on-time rates in aggregate. This situation is an example of Simpson's paradox: the trend observed in the stratified data is no longer apparent in the aggregated data. This is likely due to the extreme number of on-time Am West flights to Phoenix: 4840. This likely skewed the aggregate data to show a high percentage of on-time flights for Am West.

Overall, this assignment could be extended by incorporating more data into the analysis (more airlines, more destinations, average delay times, etc.). Additionally, if historical data were available, the on-time rates for each airline could be calculated for each year or month of operation. Lastly, this assignment could be completed using a mySQL database for storage and then SQL manipulations could be practiced.
